---
title: 'STAT/CSSS 503: Assignment 3'
author: Everett Mumford
date: "May 17, 2017"
bibliography: assignment3.bib
output:
  html_document:
    toc: false
    number_sections: true
  pdf_document:
    toc: false
    number_sections: true    
---


```{r include=FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```


```{r, message=FALSE}
library("AER")
library(tidyverse)
```

# Problem 1
```{r}
data <- read.csv("Ch09.Ex1.CivilWarRainInstrumentData/RainIV.csv")
```

## Problem 1A
```{r}
OneA <- lm(InternalConflict~LaggedGDPGrowth, data=data)
summary(OneA)
```
The coefficient on LaggedGDPGrowth denotes the change in the probability of internal conflict for each one unit increase in LaggedGDPGrowth (since InternalConflict ~ LaggedGDPGrowth is a linear probability model). Therefore, for each one unit increase in LaggedGDPGrowth, there is a .08 decrease in the probability of civil war.

## Problem 1B
```{r}
OneB <- lm(InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac, data=data)
summary(OneB)
```
The results do not establish causation. As someone mentioned in class, the coefficient on LaggedGDPGrowth is not significant and the standard error is more than twice as large as the estimated coefficient. Furthermore, there are many other omitted variables (e.g. presence of famine) that are not included on the right side of the equation.

## Problem 1c
The two conditions required for a good instrument are the inclusion and exclustion criteria. In this case, the inclusion criteria necessitates that LaggedRainfallGrowth be correlated with LaggedGDP. We can test for this by regressing LaggedGDPGrowth on LaggedRainfall and all of our other covariates and determining if the coefficient on LaggedRainfall is statistically significant. The coefficient on LaggedRainfallGrowth is statistically significant, so LaggedRainfall satisfies the inclusion criteria. The t value is greater than 3, so we can even argue that LaggedRainfallGrowth is a strong instrument.
```{r}
OneC <- lm(LaggedGDPGrowth ~ LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac, data=data)
summary(OneC)
```
The exclusion criteria stipulates that the instrument not be correlated with the dependent variable, except through it's relationship with the independent variable. There is no statistical test to determine if the exclusion criteria is satisfied (though falsification tests are sometimes done in an attempt to statistically prove that the exclusion criteria is met). Therefore, we need to conceptually decide if the exclusion is met. As someone mentioned in class, rainfall is associated with food scarcity. Food scarcity is often a harbinger of conflict. Therefore, I would argue that rainfall does not satisfy the exclusion criteria and is therefore not a suitable instrumental variable. 

## Problem 1D
If both the internal and external criteria are met, then instrumental variable analysis can generate exogenous variation in LaggedGDPGrowth and determine how that exogenous variation affects the probability of internal confict. The first stage regression produces estimates of LaggedGDPGrowth that vary only with changes in the instrument and control variables. Therefore, these estimated X values are not correlated with the error term in the second-stage regression (thus creating exogenous variation in X). As long as the instrument is not correlated with the dependent variable (and assuming that no other omitted variable bias is present in our second stage regresison), we can estimate the causal effect of LaggedGDPGrowth by regressing internal conflict on the y-hat values of LaggedGDPGrowth from the first stage regression.

## Problem 1E
```{r}
OneE <- ivreg(InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac | LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac, data=data)
summary(OneE)
```
The coefficient on LaggedGDPGrowth denotes a 206% decrease in the probability of conflict for each one unit increase in LaggedGDPGrowth. The standard error is quite large (1.84).

## Problem 1f
```{r}
OneF_1stStage <- lm(LaggedGDPGrowth ~ LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_name, data=data)
summary(OneF_1stStage)
```
In the first stage, we can see that the inclusion criteria is satisfied (the coefficient on LaggedRainfallGrowth is statistically significant) and LaggedRainfallGrowth is a strong instrument (the t statistic greater than 3).

```{r}
OneF <- ivreg(InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_name | LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_name, data=data)
summary(OneF)
```
By including country fixed effects, we're effectively comparing the effect of LaggedGDPGrowth on the probability of internal conflict within each country. The coefficient on LaggedGDPGrowth denotes a 285.34% decrease in civil conflict for each one unit increase in LaggedGDPGrowth.

## Problem 1g
```{r}
OneG <- lm(LaggedGDPGrowth ~ LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_name, data=data)
residuals <- residuals(OneG)
OLS <- lm(InternalConflict ~ I(LaggedGDPGrowth - residuals) + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_name, data=data)
summary(OLS)
```
The coefficient on LaggedGDPGrowth is the same in the OLS and 2SLS models (-2.85). In the 2SLS model, the y-hat values for LaggedGDPGrowth from the first stage are used as the independent variable in the second stage regression. The 2SLS and OLS specifications are related. The y-hat values from the first stage of the 2SLS model are simply the fitted values of the first stage regression; the fitted values are equal to the actual values minus a residual. So subtracting the residuals from the actual values in the OLS model yields the same coefficient on LaggedGDPGrowth, since it makes the independent variables equivalent in each model. 
Endogeneity is being handled in the OLS specification for the same reason it's handled in the 2SLS model; the independent variable is now a product of the effect of the instrument and other independent variables on economic growth. Therefore changes in economic growth are not correlated with the error term when regressing conflict on growth.

# Problem 2
```{r}
tv.data <- read.csv("Ch09.Ex2.TelevisionExperimentData/news_study_MAB.csv")
```

## Problem 2a
```{r}
TwoA <- lm(InformationLevel ~ WatchProgram, data=tv.data)
summary(TwoA)
```
The coefficient on WatchProgramYes denotes a .3 increase in the mean of InformationLevel for those that watched the program versus those that did not. The results could certainly be biased by omitted variables. For instance, control variables included in part 2a are certainly leading to omitted variable bias in the bivariate OLS. 

## Problem 2b
```{r}
TwoB <- lm(InformationLevel~WatchProgram + PoliticalInterest + ReadNews + Education, data=tv.data)
summary(TwoB)
```
The estimated coefficient is lower in the second regression, as is the standard error. Both coefficients are statistically significant. I would argue that we have not defeated endogeneity. There are plenty of other omitted variables that could be biasing the results (e.g. variables that may determine someone's a priori opinion on affirmative action). Therefore, I would argue that the equation does not estimate an exogenous effect of X on Y.

## Problem 2c
The assignment variable will be a good instrument if it satisfies the inclusion and exclusion criteria. The inclusion criteria in this case stipulates that the treatment is correlated with the level of information that someone has on the topic. We can test this quite easily be regressing WatchProgram on TreatmentGroup and determining if the coefficient is statistically significant. The coefficient is statistically significant and is well above 3, thus we have a very strong instrument.
```{r}
tv.data$TreatmentGroup <- as.character(tv.data$TreatmentGroup)
tv.data$TreatmentGroup[tv.data$TreatmentGroup != "0"] <- "1"

# Referenced this StackOverflow post while writing line of code above -- "Replace a Value in a Data Frame Based on a Conditional (`if`) Statement in R." StackOverflow. Stack Exchange, n.d. Web. 23 May 2017. <https://stackoverflow.com/questions/5824173/replace-a-value-in-a-data-frame-based-on-a-conditional-if-statement-in-r>. question posted on 4/28/11. referenced the accepted answer posted on 4/28/11.

tv.data$TreatmentGroup <- as.factor(tv.data$TreatmentGroup)
tv.data$WatchProgram <- as.numeric(tv.data$WatchProgram)
TwoC <- lm(WatchProgram ~ TreatmentGroup + PoliticalInterest + ReadNews + Education, data=tv.data)
summary(TwoC)
```

There is no test for the exclusion criteria. We need to determine if TreatmentGroup is correlated with InformationLevel, outside of through WatchProgram. Conceptually, it seems reasonable to argue that the exclusion criteria is satisfied.

## Problem 2d
```{r}
TwoD <- ivreg(InformationLevel ~ WatchProgram + PoliticalInterest + ReadNews + Education | TreatmentGroup + PoliticalInterest + ReadNews + Education, data=tv.data)
summary(TwoD)
nrow(model.frame(TwoD))
```

```{r}
summary(TwoB)
nrow(model.frame(TwoB))
```
The coefficients are quite similar in the two regressions. The standard error is quite a bit higher in 2d, and thus the coefficient on WatchProgramYes is statistically significant at the alpha=.1 level in 2B but not 2D. The sample sizes and degrees of freedom in each regression are the same (498 and 472, respectively).

## Problem 2e
I am more confident in the results of the 2SLS model than the regression run in 2B. Both the exclusion and inclusion criteria appear to be met. I am more confident in the 2SLS model because of all of the endogeneity in the OLS model.


# Problem 4
```{r}
inmates.data<- read.csv("Ch09.Ex4.EducationCrimeData/inmates.csv")
```

## Problem 4a
```{r}
FourA <- lm(prison ~ educ + age + AfAm + state + year, data=inmates.data)
summary(FourA)
```
Since we're estimating a linear probability model, the coefficient denotes how the probability of being prison changes with respect to a one year increase in an indivdual's education. In the case of this model, each extra year of education is associated with a .001 decrease in the probability of being in prison. The coefficient is statistically significant (though perhaps not substantively significant) in this model.

## Problem 4b
We cannot conclude that increasing education will decrease crime rates based on this model alone. For starters, there are several omitted variables that are not included in the equation that are leading to endogeneity (e.g. household income). In addition, there may be some reverse causality, thus making it difficuly to use a simple OLS equation to determine causality.

## Problem 4c
The inclusion criteria stipulates that the instruments have a significant effect on education. We can use an F-test to determine if all of the variables satisfy the inclusion criteria. The null hypothesis of the F test is that the coefficient on each instrument is zero. Therefore, we'll reject the null if any coefficient is statistically significant (that is, if any instrumental variable satisfies the inclusion critera). Since the F-statistic is statistically significant and greater than 10, the inclusion criteria is satisfied (at least one of the package of instrumental variables is significantly correlated with education). 
```{r}
unrestricted_stage1 <- lm(educ ~ age + AfAm + state + year + ca9 + ca10 + ca11, data=inmates.data)
restricted_stage1 <- lm(educ ~ age + AfAm + state + year, data=inmates.data)
anova(unrestricted_stage1, restricted_stage1)
```
The exclusion criteria stipulates that compulsory attendance laws only affect the probability of being in prison through the effect of the law on years of education. There is no statistical test to determine if the exclusion criteria is satisfied, but conceptually it seems reasonable to argue that the effect of compulsory education laws only affect probability of imprisonment through number of years of education. Therefore, I would argue that the compulsory education laws satisfy both the inclusion and exclusion criteria.

## Problem 4d
```{r}
FourD <- ivreg(prison ~ educ + age + AfAm + state + year | ca9 + ca10 + ca11 + age + AfAm + state + year, data=inmates.data)
summary(FourD)
```
Since the second stage equation is a linear probability model, the coefficient on education denotes how the probability of being in prison changes with each one year increase in education levels. In the case of this model, a one year increase in education leads to a .002 decrease in the probability of ending up in prison.

## Problem 4e
The standard error is larger in the 2SLS regression than in the OLS regression. I do not think that the larger standard error is necessarily a problem. All else being equal, a larger standard error necessarily means that we are more uncertain about the effect of our independent variable on our dependent variable. In the case of the 2SLS and OLS regressions, our effect size is so small (.002 and .001 respectively) that even a small standard error could cause our confidence interval to cross zero, thus causing us to be to be unsure if the independent variable has a negative, null, or positive effect on the dependent variable. In both the OLS and 2SLS regressions, the standard errors are small enough to lead us to believe that increasing education leads to a decrease in the probability of imprisonment.


# Problem 5
```{r}
econ_data <- read.csv("Ch09.Ex5.GrowthDemocracyData/democracy_income.csv")
```

## Problem 5a
```{r}
FiveA <- lm(democracy_fh ~ lag(log_gdp), data=econ_data)
summary(FiveA)
```
The results indicate that a one percent increase in last year's GDP leads a .21 higher score on the freedom house measure.

There are several concerns. For instance, there are many omitted variables on the right side of the equation (e.g. a measure of current conflict in each country). Furthermore, it's not clear to me whether a one year lag on GDP is appropriate. I'm not sure if the lag should be longer or if there should be any lag at all. As was mentioned in the question, there may be some reverse causality, leading to plenty of endogeneity in the simplified bivariate OLS above. In addition, the model intercept (-1.22) is non-sensical, since the freedom house index ranges from 0 to 1.


## Problem 5b
```{r}
FiveB <- lm(democracy_fh ~ lag(log_gdp) + year + CountryCode, data=econ_data)
summary(FiveB)
```
The coeffiient on lagged,logged GDP is very similar in the fixed-effects model (.22) as it was in the bivariate OLS (.21). The intercept changes quite a bit; it was -1.22 in the bivariate OLS and is 3.6 in the fixed effects model.


## Problem 5c
The inclusion criteria stipulates that world income be correlated with logged GDP for a given country. We can test whether the inclusion criteria is satisfied statistically. Since the coefficient on worldincome is statistically significant when we regress log_gdp on worldincome, we can say that the inclusion criteria is satisfied. Since the t-statistic is greater than 3, we can say that worldincome is a strong instrument. 
```{r}
FiveC <- lm(log_gdp ~ worldincome + year + CountryCode, data=econ_data)
summary(FiveC)
```
The exclusion criteria stipulates that world income not be correlated with democracy, except through a country's GDP. There is no statistical test to determine whether the exclusion criteria is satisfied. I would argue that the exclusion criteria is not satisfied, however. Conceptually, it seems entirely probable to me that the income of trading partner countries is correlated with violence in trading partner countries. That violence can affect a country's freedom house score in many ways. For instance, perhaps violence in a neighboring country will cause a mass exodus of people from the country experiencing violence into that country's trading partners. An inflow of refugess can change the political climate in a country markedly.


## Problem 5d
```{r}
FiveD <- ivreg(democracy_fh ~ lag(log_gdp) + year + CountryCode | lag(worldincome) + year + CountryCode, data=econ_data)
summary(FiveD)
```
The coefficient is quite similar in the 2SLS, OLS, and fixed effects models (.19, .21, and .22, respectively). The standard errors are quite a bit smaller in the fixed effects model (.008) and OLS model (.008) than in the 2SLS model (.04), as is to be expected.

# References {-}
"Replace a Value in a Data Frame Based on a Conditional (`if`) Statement in R." StackOverflow. Stack Exchange, n.d. Web. 23 May 2017. <https://stackoverflow.com/questions/5824173/replace-a-value-in-a-data-frame-based-on-a-conditional-if-statement-in-r>. question posted on 4/28/11. referenced the accepted answer posted on 4/28/11.
