---
title: 'STAT/CSSS 503: Assignment 3'
author: Jeffrey Arnold
date: "May 17, 2017"
bibliography: assignment3.bib
output:
  html_document:
    toc: false
    number_sections: true
  pdf_document:
    toc: false
    number_sections: true    
---

```{r include=FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```


# Instructions {-}

This repository contains the assignment instructions.
Submitted solutions will use a **separate** repository.

1. Fork [UW-CSSS-503/assignment-2017-3-submissions](https://github.com/UW-CSSS-503/assignment-2017-3-submissions) repository to your account.
2. Edit the file `assignment-3-solutions.Rmd` with your solutions to the problems.
3. Submit a pull request to have it graded. Include either or both a HTML and PDF file.

This assignment will require the following R packages:
```{r, message=FALSE}
library("AER")
```

# Problem 1

@Bailey2016a Ex 9.1

# Problem 2

@Bailey2016a Ex 9.2

# Problem 3

@Bailey2016a Ex 9.4

# Problem 4

@Bailey2016a Ex 9.5

# References {-}
